---
title: "Type I Error in Bootstrapping"
author: "Dan Hodges"
date: "https://github.com/ST541-Fall2018/hodgesda-project-comparing-typeI-bootstrap"
output:
  ioslides_presentation: default
  beamer_presentation: default
  slidy_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
library(ggplot2)
presentdata <- readRDS("/cloud/project/results/presentdata.rds")
pepper <- readRDS("/cloud/project/results/pop1.rds")
pepper.data <- data.frame(pepper)
ibo <- readRDS("/cloud/project/results/pop2.rds")
ibo.data <- data.frame(ibo)
```

## Motivation
- Bootstrapping is taught as an alternative to traditional testing methods.
- It can seem "magical" (One procedure to test any statistic).
- Statistics faculty caution against "Teaching students enough to be dangerous."

## Motivation
- How does the bootstrapping procedure compare to a more traditional test in terms of Type I error?
- This project will compare bootstrapping the mean to the T-test.

## Quick Overview
- Simulated two populations: one symmetric, one not.
- For a vector of sample sizes perform many* bootstraps and estimate the Type-I error rate, alpha.

## Meet the Populations

```{r}
ggplot(pepper.data,aes(x=pepper))+geom_histogram(bins=100)+
  xlab("x")+theme_bw()
```

## Meet the Populations

```{r}
ggplot(ibo.data,aes(x=ibo))+geom_histogram(bins=100)+
  xlab("x")+theme_bw()
```

## Preliminary Results

```{r}
ggplot(presentdata[presentdata$population=="Pepper",], aes(x=as.factor(n), y=alpha, color=as.factor(nboot), group=as.factor(nboot))) +
  geom_point() +
  geom_line()+
  theme_bw()+
  ggtitle("Estimated Type I Error from Symmetric Population")+
  xlab("Sample Size")+
  ylab("Alpha")+
  scale_colour_discrete(name="Bootstrap Size")

```

## Preliminary Results

```{r}
ggplot(presentdata[presentdata$population=="Ibo",], aes(x=as.factor(n), y=alpha, color=as.factor(nboot), group=as.factor(nboot))) +
  geom_point() +
  geom_line()+
  theme_bw()+
  ggtitle("Estimated Type I Error from Asymmetric Population")+
  xlab("Sample Size")+
  ylab("Alpha")+
  scale_colour_discrete(name="Bootstrap Size")
```

## Thank You

Github Repo: https://github.com/ST541-Fall2018/hodgesda-project-comparing-typeI-bootstrap